---
title: "Low-Cost PM Sensor Tutorial"
author: "Dr. Naomi Zimmerman"
date: "1/12/2021"
output: 
  html_document:
    toc: true
bibliography: packages.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.width = 7, 
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

## PM~2.5~ Assessment Tutorial
In this tutorial, we will download some PurpleAir data from sensors installed at the Burnaby South Regulatory Monitoring Station in Vancouver, Canada. We will take advantage of the [@R-AirSensor] package
For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:




```{r packages}
if (!require(installr)) install.packages('installr')
library(installr)

if (!require(stringr)) install.packages('stringr')
library(stringr)

if (!require(chron)) install.packages('chron')
library(chron)

if (!require(prodlim)) install.packages('prodlim')
library(prodlim)

if (!require(gtools)) install.packages('gtools')
library(gtools)

if (!require(AirSensor)) install.packages('AirSensor')
library(AirSensor)

if (!require(dplyr)) install.packages('dplyr')
library(dplyr)

if (!require(ggplot2)) install.packages('ggplot2')
library(ggplot2)
```

## Set up a local directory

To download the PurpleAir data in this tutorial, we will take advantage of the [@R-AirSensor] package. The AirSensor package needs to know where processed data will live. For this report, we will specify a local archiveBaseDir where downloaded and processed data will live. The following code specifies a local directory archive, checks for its existence and creates it if needed. It also checks to see if you have downloaded the required spatial data sets (to extract PurpleAir data by Country Code) and installs them if necessary.

```{r set_up}

## Create a folder in your home directory for this tutorial called 'PM_Sensor_Tutorial'
new_wd <- path.expand("~/PM_Sensor_Tutorial")
if ( !dir.exists(new_wd) ) {
  dir.create(new_wd)
}
setwd(new_wd)

## Create a folder in 'PM_Sensor_Tutorial' for storing the downloaded PurpleAir data.
archiveBaseDir <- path.expand("~/PM_Sensor_Tutorial/Data")
if ( !dir.exists(archiveBaseDir) ) {
  dir.create(archiveBaseDir)
}
setArchiveBaseDir(archiveBaseDir)

## Checking to see if you have the required spatial data, and installing it un a sub-folder of PM_Sensor_Tutorial/Data/Spatial if needed
if (!require(MazamaSpatialUtils)) install.packages('MazamaSpatialUtils')
library(MazamaSpatialUtils)

if ( !dir.exists('~/PM_Sensor_Tutorial/Data/Spatial') ) {
  dir.create('~/PM_Sensor_Tutorial/Data/Spatial')
  setSpatialDataDir('~/PM_Sensor_Tutorial/Data/Spatial')
  installSpatialData()
  installSpatialData('NaturalEarthAdm1')
}
```
## Load synoptic data for Canada

We will use the AirSensor pas_createNew() function to create a Purple Air Synoptic (pas) object containing all the spatial metadata associated with purple air monitors in Canada (we will call this object 'pas_CAN'). See: [https://mazamascience.github.io/AirSensor/articles/pas_introduction.html] for more details.

The following code checks to make sure the pas_CAN file exists in your archiveBaseDir directory and, if missing, recreates it.

```{r data_download}
filePath_pas <- file.path(archiveBaseDir, "pas_CAN.rda")

if ( !file.exists(filePath_pas) ) {
  initializeMazamaSpatialUtils(spatialDataDir = "~/PM_Sensor_Tutorial/Data/Spatial")
  pas_CAN <- pas_createNew(countryCodes = "CA", includePWFSL = TRUE, lookbackDays = 1,baseUrl = "https://www.purpleair.com/json?all=true")
  save(pas_CAN, file = filePath_pas)
}

pas_CAN <- get(load(filePath_pas))
```

## Visualize snapshot of Air Quality in Canada
```{r map}
pas_leaflet(pas_CAN)
```

## Download data from a Vancouver PurpleAir (co-located with Regulatory Instruments)

Metro Vancouver, the local regulatory agency, has two PurpleAir monitors co-located with their 'Burnaby South' Regulatory Monitoring Site [https://gis.metrovancouver.org/maps/Air]. This is a good site for our tutorial, as there is downloadable regulatory data and low-cost sensor data available. The following code snippet 

```{r time_series_download}
filePath_Burnaby <- file.path(archiveBaseDir, "pat_Burnaby.rda")

if ( file.exists(filePath_Burnaby) ) {
  pat_Burnaby <- get(load(filePath_Burnaby))
} else {
  pat_Burnaby <- pat_createNew(
    pas = pas_CAN, 
    label = "NAPSID: 100119 - 1", 
    startdate = 20200901, 
    enddate = 20201031
  )
  save(pat_Burnaby, file = filePath_Burnaby)
}
```

## Including Plots

You can also embed plots, for example:

```{r map2, echo=FALSE}
lon <- pat_Burnaby$meta$longitude
lat <- pat_Burnaby$meta$latitude
pas_vancouver <- 
  pas_CAN %>%
  pas_filterNear(
    longitude = lon, 
    latitude = lat, 
    radius = "25 km"
  ) 
pas_leaflet(pas_vancouver)
```
## Extract the data
```{r data}
data_Burnaby <- pat_extractData(pat_Burnaby)
head(data_Burnaby)

validation <- pat_externalFit(pat_Burnaby)

ref_data <- validation[["model"]][["both_data$pwfsl_pm25"]]
lcs_data <- validation[["model"]][["both_data$pa_pm25"]]

mean_lcs = mean(lcs_data) #verified
mean_ref = mean(ref_data) # verified
bias = mean_lcs - mean_ref #verified
sd_data = sqrt(var(lcs_data)) #verified
r = cor(lcs_data,ref_data)  #verified
          
error = lcs_data - ref_data  #verified
error_unbiased = error - bias #verified
error_fractional = error/ref_data #verified
error_fractional[ref_data == 0] = NA #verified
          
RMSE = sqrt(mean(error^2)) #verified
MAE = mean(abs(error)) #verified
            
RMSE_unbiased = sqrt(mean(error_unbiased^2)) #verified
MAE_unbiased =  mean(abs(error_unbiased)) #verified
            
MBE = mean(error_fractional,na.rm = TRUE) #verified
CRMSE = sqrt(mean(error_fractional^2,na.rm = TRUE)) #verified


# EPA Error Statistics
ref_data_LDL = ref_data
ref_data_LDL[ref_data_LDL<2]=NaN

lcs_data_LDL = lcs_data
lcs_data_LDL[lcs_data_LDL<2]=NaN

LDL_index = which((!is.na(lcs_data_LDL)) & (!is.na(ref_data_LDL)))
      if (!is.empty(LDL_index)) {
            avg_data_LDL = (lcs_data_LDL + ref_data_LDL)/2
            error_pct = 100*((lcs_data_LDL- ref_data_LDL)/avg_data_LDL)
            error_pct[is.infinite(error_pct)] = NA
            LDL_index_a = which(!is.na(error_pct))
            error_pct = error_pct[LDL_index_a]
            
            d_25p = quantile(error_pct,0.25) #verified
            d_75p = quantile(error_pct,0.75) #verified
            
            error_pct_squared = error_pct^2 #verified
            error_pct_absolute = abs(error_pct) #verified
            
            n_data = length(LDL_index_a) #verified
            
            error_pct_var = var(error_pct) #verified
            error_pct_std = sqrt(error_pct_var) #verified
             
            error_pct_squared_var = var(error_pct_squared) #verified
            error_pct_squared_std = sqrt(error_pct_squared_var) #verified
            
            error_pct_sum = sum(error_pct) #verified
            error_pct_absolute_sum = sum(error_pct_absolute) #verified
            error_pct_squared_sum = sum(error_pct_squared) #verified
            
            AB = mean(error_pct_absolute) #verified
            AS = sqrt(((n_data*error_pct_squared_sum)-(error_pct_absolute_sum^2))/(n_data*(n_data-1)))
              
            CV_Score = (sqrt(((n_data*error_pct_squared_sum)-(error_pct_sum^2))/(n_data*(n_data-1)))*sqrt((n_data-1)/(qchisq(0.1,(n_data-1)))))/100
            
            Bias_Score = (AB + ((qt(0.95,(n_data-1)))*(AS/sqrt(n_data))))/100
            
          } else {
            CV_Score = NA
            Bias_Score = NA
          }


```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## References
```{r, include=FALSE}
knitr::write_bib(c(.packages()), "packages.bib")
```
