---
title: "Low-Cost PM Sensor Tutorial"
author: "Dr. Naomi Zimmerman"
date: "1/12/2021"
output: 
  html_document:
    toc: true
bibliography: packages.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.width = 7, 
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

## PM~2.5~ Assessment Tutorial
In this tutorial, we will download some PurpleAir data from sensors installed at the Burnaby South Regulatory Monitoring Station in Vancouver, Canada. We will take advantage of the [@R-AirSensor] package
For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r packages}
library(installr)
library(stringr)
library(chron)
library(prodlim)
library(gtools)
library(AirSensor)
library(dplyr)
library(ggplot2)
library(MazamaSpatialUtils)

setSpatialDataDir('~/Data/Spatial')
loadSpatialData("NaturalEarthAdm1")
initializeMazamaSpatialUtils()

pas_CA <- pas_createNew(countryCodes = "CA",includePWFSL = TRUE,lookbackDays = 1,baseUrl = "https://www.purpleair.com/json?all=true")

pat_Burnaby <- pat_createNew(
  pas = pas_CA, 
  label = "NAPSID: 100119 - 1",
  startdate=20200901,
  enddate=20201231)
```

## Including Plots

You can also embed plots, for example:

```{r map, echo=FALSE}
lon <- pat_Burnaby$meta$longitude
lat <- pat_Burnaby$meta$latitude
pas_vancouver <- 
  pas_CA %>%
  pas_filterNear(
    longitude = lon, 
    latitude = lat, 
    radius = "25 km"
  ) 
pas_leaflet(pas_vancouver)
```
## Extract the data
```{r data}
data_Burnaby <- pat_extractData(pat_Burnaby)
head(data_Burnaby)

validation <- pat_externalFit(pat_Burnaby)

ref_data <- validation[["model"]][["both_data$pwfsl_pm25"]]
lcs_data <- validation[["model"]][["both_data$pa_pm25"]]

mean_lcs = mean(lcs_data) #verified
mean_ref = mean(ref_data) # verified
bias = mean_lcs - mean_ref #verified
sd_data = sqrt(var(lcs_data)) #verified
r = cor(lcs_data,ref_data)  #verified
          
error = lcs_data - ref_data  #verified
error_unbiased = error - bias #verified
error_fractional = error/ref_data #verified
error_fractional[ref_data == 0] = NA #verified
          
RMSE = sqrt(mean(error^2)) #verified
MAE = mean(abs(error)) #verified
            
RMSE_unbiased = sqrt(mean(error_unbiased^2)) #verified
MAE_unbiased =  mean(abs(error_unbiased)) #verified
            
MBE = mean(error_fractional,na.rm = TRUE) #verified
CRMSE = sqrt(mean(error_fractional^2,na.rm = TRUE)) #verified


# EPA Error Statistics
ref_data_LDL = ref_data
ref_data_LDL[ref_data_LDL<2]=NaN

lcs_data_LDL = lcs_data
lcs_data_LDL[lcs_data_LDL<2]=NaN

LDL_index = which((!is.na(lcs_data_LDL)) & (!is.na(ref_data_LDL)))
      if (!is.empty(LDL_index)) {
            avg_data_LDL = (lcs_data_LDL + ref_data_LDL)/2
            error_pct = 100*((lcs_data_LDL- ref_data_LDL)/avg_data_LDL)
            error_pct[is.infinite(error_pct)] = NA
            LDL_index_a = which(!is.na(error_pct))
            error_pct = error_pct[LDL_index_a]
            
            d_25p = quantile(error_pct,0.25) #verified
            d_75p = quantile(error_pct,0.75) #verified
            
            error_pct_squared = error_pct^2 #verified
            error_pct_absolute = abs(error_pct) #verified
            
            n_data = length(LDL_index_a) #verified
            
            error_pct_var = var(error_pct) #verified
            error_pct_std = sqrt(error_pct_var) #verified
             
            error_pct_squared_var = var(error_pct_squared) #verified
            error_pct_squared_std = sqrt(error_pct_squared_var) #verified
            
            error_pct_sum = sum(error_pct) #verified
            error_pct_absolute_sum = sum(error_pct_absolute) #verified
            error_pct_squared_sum = sum(error_pct_squared) #verified
            
            AB = mean(error_pct_absolute) #verified
            AS = sqrt(((n_data*error_pct_squared_sum)-(error_pct_absolute_sum^2))/(n_data*(n_data-1)))
              
            CV_Score = (sqrt(((n_data*error_pct_squared_sum)-(error_pct_sum^2))/(n_data*(n_data-1)))*sqrt((n_data-1)/(qchisq(0.1,(n_data-1)))))/100
            
            Bias_Score = (AB + ((qt(0.95,(n_data-1)))*(AS/sqrt(n_data))))/100
            
          } else {
            CV_Score = NA
            Bias_Score = NA
          }


```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## References
```{r, include=FALSE}
knitr::write_bib(c(.packages()), "packages.bib")
```
